Subset coercions in \Coq

### 2006-05-01 papers/article/main.tex
Algo typing

\documentclass[twocolumn]{article}
\usepackage{abbrevs}
\usepackage{array}
%\usepackage{
\usepackage{amsmath}

@@

\subsubsection{The typing judgment}
\Russell{}'s typing system (figure \vref{fig:typing-decl-rules}) is based
on the \CCfull \cite{Barras}, with sigma types but without universes.
This restriction may be removed in future work, but cause no problem for
our main purpose which is programming. The judgement $`G \typed t : T$
means $t$ is a well-typed term of type $T$ in environment $`G$, and $`G
\typed T \subd U$ means $T$ is convertible to $U$ in environment $`G$.
Following the presentation as a pure type system \cite{PTS}, 
the set of sorts $\PTSsorts$ is defined as $\setproptype$, the
axioms are $\PTSaxioms = \{ (\Set, \Type), (\Prop, \Type) \}$ and the
rules $\PTSrules$ are defined by the following (functional) relation:
\axiomsd

@@

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgment. It is well-known \cite{Luo, Pottinger?} that $\beta\pi$ is
confluent only on well typed terms, so we have to ensure that our rules,
when read from bottom to top, keep well-typedness. That's the purpose of
the $`G \typed U : s$ premise of the \irule{SubTrans} rule. This rules
ensures that our judgement constructs an \emph{equivalence} and has
proper modularity. The next two rules do context closure for dependent
products and sums. It is remarkable that we use contravariance for
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgements because we have symetry. However, it will
become important when we create coercions (see
\vref{fig:subtyping-impl-rules}).

@@

\subsubsection{Properties}
\begin{paragraph}{Assumptions}
  We have proved basic metatheoretical results for this system in \Coq{}
  \cite{CCP}. We have assumed that it is strongly normalizing (SN) and has the
  following subject reduction (SR) property: if $`G \typed t : T$ and $t
  "->"_{\beta\pi} t'$ then $`G \typed t' : T' \subd T$. Strong normalization
  could be proved using the generic proof method developed in
  \cite{Geuvers} but we leave it for future work. Gang Chen
  \cite{GangChen} has studied various type systems from the
  $\lambda$-cube extended by subtyping or coercive subtyping \cite{Luo},
  including the Calculus of Constructions, and proved such results as
  SN and SR for them. It requires a very careful analysis of the system 
  to avoid cycles in the proof due to the presence of dependent types and
  conversion ; hence we also leave it for future work.
\end{paragraph}

\begin{paragraph}{Results}
  - Weakening
  - Substitution
\end{paragraph}

\subsubsection{Algorithmic typing}
  - Standard transformation for subtyping
  - mu
  - Elimination of transitivity
  - Soundness and completeness

>>>>

\documentclass[twocolumn]{article}
\usepackage{abbrevs}
\usepackage{array}
%\usepackage{
\usepackage{amsmath}
\usepackage{amsthm}

@@

\subsubsection{The typing judgment}
\Russell{}'s typing system (figure \vref{fig:typing-decl-rules}) is based
on the \CCfull \cite{Barras99,Luo90}, with sigma types but without universes.
This restriction may be removed in future work, but cause no problem for
our main purpose which is programming. The judgement $`G \typed t : T$
means $t$ is a well-typed term of type $T$ in environment $`G$, and $`G
\typed T \subd U$ means $T$ is convertible to $U$ in environment $`G$.
Following the presentation as a pure type system \cite{PTS}, 
the set of sorts $\PTSsorts$ is defined as $\setproptype$, the
axioms are $\PTSaxioms = \{ (\Set, \Type), (\Prop, \Type) \}$ and the
rules $\PTSrules$ are defined by the following (functional) relation:
\axiomsd

@@

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgment. It is well-known \cite{Luo, Pottinger?} that $\beta\pi$ is
confluent only on well typed terms, so we have to ensure that our rules,
when read from bottom to top, keep well-typedness. That's the purpose of
the $`G \typed U : s$ premise of the \irule{SubTrans} rule. This rules
ensures that our judgement constructs an \emph{equivalence} and has
proper modularity. The next two rules do context closure for dependent
products and sums. It is remarkable that we use contravariance for
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS \cite{journals/toplas/Castagna95}. It is accessory here, as we could have used \emph{co}variance and
still get the same judgements because we have symetry. However, it will
become important when we create coercions (see
\vref{fig:subtyping-impl-rules}).

@@

\subsubsection{Properties}
\begin{paragraph}{Assumptions}
  We have proved basic metatheoretical results for this system in \Coq{}
  \cite{CCP}. We have assumed that it is strongly normalizing (SN) and has the
  following subject reduction (SR) property: if $`G \typed t : T$ and $t
  "->"_{\beta\pi} t'$ then $`G \typed t' : T$. Strong normalization
  could be proved using the generic proof method developed in
  \cite{Geuvers} but we leave it for future work. Gang Chen
  \cite{GangChen} has studied various type systems from the
  $\lambda$-cube extended by subtyping or coercive subtyping \cite{Luo},
  including the Calculus of Constructions, and proved such results as
  SN and SR for them. It requires a very careful analysis of the system 
  to avoid cycles in the proof due to the presence of dependent types and
  conversion ; hence we also leave it for future work.
\end{paragraph}

We write $`G \typed T, U : s$ for $`G \typed T : s$ and $`G \typed U :
s$.

As all functionnal PTS, \Russell{} has the unicity of sorting property,
hence we can prove:
\begin{lemma}[Well-sortedness of equivalence derivations]
  If $`G \typed T : s_1$, $`G \typed U : s_2$ and $`G \typed T \sub U$
  then $`G \typed T, U : s$ and $s = s_1 = s_2$.
\end{lemma}
\begin{proof}
  We will use the abbreviation $`G \typed T \sub U : s$ for $`G \typed T,
  U : s$ and $`G \typed T \sub U$ in the following.

  By induction on the subtyping derivation:
  \case{SubConv} We have $T \eqbp U$, we apply unicity of sorting.

  \case{SubTrans} We have by induction $`G \typed T \sub S : s$ and 
  $`G \typed S \sub U : s$. Hence we have $`G \typed T, U : s$.

  \caseother{} Other cases are straightforward.
\end{proof}

We will omit $`G$ where unnecessary.
The essential properties of the $\_ \typed \_ \subd \_$ judgment are 
its reflexivity ($\forall A, A \subd A$), symmetry ($\forall A B, A \subd
B "<=>" B \subd A$) and transitivity by \irule{SubTrans}.

As typing and equivalence are mutually recursive, we will generaly prove
properties by mutual induction on the two derivations.
Our system has good structural properties like weakening, thinning and
substitution.

\begin{lemma}[Weakening]
  Suppose $\typedwf `G, x : S, `D$.
  If $`G, `D \typed t : T$ then $`G, x : S, `D \typed t : T$. 
  If $`G, `D \typed T \sub U : s$ then $`G, x : S, `D \typed T \sub U : s$.
\end{lemma}
\begin{proof}
  Straightforward mutual induction.
\end{proof}

Similarly for thining and substitution. An important property for
dependent type systems is to have invariance of judgments which are
definitionaly equal (here equivalent). Hence the following lemma:
\begin{lemma}
  Suppose $`G \typed U \sub V : s$.
  If $`G, x : V, `D \typed t : T$ then $`G, x : U, `D \typed t : T$.
  If $`G, x : V, `D \typed S \sub T : s'$ then
  $`G, x : U, `D \typed S \sub T : s'$.
\end{lemma}
\begin{proof}
  By mutual induction on the typing and equivalence derivations.
\end{proof}

We stop here about the theoretical properties of this first system, as
it gives no new insights for our purpose and we will focus more on the
algorithmic system's properties.

\subsubsection{Algorithmic typing}
We use a standard transformation to get an algorithm from our typing
rules, basically restricting the conversion (or subsumption) rule
(\irule{Coerce}) to the premises of introduction and elimination rules.
\TODO{App?, SumDep?, Pi?}. Our algorithmic judgement $`G \typea t : T$
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. This notion is presented in \PVS{} by
means of a $\mu_0$ operator which was readily adaptable in our setting
\ref{fig:mualgo}. In Gang Chen's work, the notion is formalized by a
new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator.

The next thing to do is to construct a decidable judgement for
equivalence given two well-sorted types. We denote it by $`G \typea T
\suba U$ (figure \ref{fig:subta-rules}).
In our case, we have to check that conversion is decidable, which is the
case. We also need to eliminate the transitivity rule which is not
syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Elimination of transitivity]
  If $`G \typea T \suba U$ and $`G \typea U \suba V$, then $`G \typea T
  \suba V$.
\end{theorem}
\begin{proof}
  By lexicographic induction on the pair of derivations.
  \TODO{example}
\end{proof}

Classicaly, we have the following correspondence between the two
systems:
\begin{theorem}[Soundness]
  If $`G \typea t : T$ then $`G \typed t : T$.
  If $`G \typea T \suba U$ then $`G \typed T \subd U$.
\end{theorem}

\begin{theorem}[Completeness]
  If $`G \typed t : T$ then $`G \typea t : T' \suba T$.
  If $`G \typed T \subd U$ then $`G \typea T \suba U$.
\end{theorem}

Here is the desired property of the algorithmic system:
\begin{theorem}[Decidability of typing and equivalence]
  $`E T, `G \typea t : T$ and $`G \typea T \suba U$ are decidable
  problems.
\end{theorem}
